---
title: "Stats506_PS6"
author: "Heleyna Tucker"
format: html
editor: visual
---

# Problem 1 - Plotting

Using the "nnmaps" data set again. ggplot2 is recommending for these.

### Part A: Produce a graph of the mean monthly temp in Celsius

Produce a graph of the **mean** monthly temperature in **celsius**. Draw a scatter-plot where the x-axis is month and the y-axis is the average monthly temperature in celsius. Add a line connecting the points within each season and color the lines and points by season (same color for lines and points in the same season). Be sure both x-axis and the legend are ordered appropriately, and that all labels/titles are clear and ready for publication (e.g. no variable names).

Read in the data and visualize below:

```{r}
library(tidyverse)
nnmaps <- read_csv('/Users/19892/OneDrive/Documents/STATS506/ProblemSets/Stats506_PS5/chicago-nmmaps (1).csv')
head(nnmaps)
```

Convert Fahrenheit and add a column to the dataset with this conversion:

```{r}
# Function to convert Fahrenheit to Celsius
fahrenheit_to_celsius <- function(fahrenheit) {
  celsius <- (fahrenheit - 32) * (5/9)
  return(celsius)
}

nnmaps$tempC <- fahrenheit_to_celsius(nnmaps$temp)
head(nnmaps)
nnmaps$month
```

```{r}
library(dplyr)
# Calculate the average temperature for each month
avg_temp_per_month <- tapply(nnmaps$tempC, nnmaps$month_numeric, mean)


# Create a new data frame with average temperatures and seasons
avg_temp_df <- data.frame(month_num = names(avg_temp_per_month), avg_temp = avg_temp_per_month)
avg_temp_df$month <- month.abb[as.numeric(avg_temp_df$month_num)]

avg_temp_df <- avg_temp_df %>%
  mutate(season = case_when(
  month %in% c("Jan", "Feb", "Mar") ~ "Winter",
  month %in% c("Apr", "May", "Jun") ~ "Spring",
  month %in% c("Jul", "Aug", "Sep") ~ "Summer",
  month %in% c("Oct", "Nov", "Dec") ~ "Autumn"))
avg_temp_df
```

```{r}
library(ggplot2)

# Order the "Month" factor
avg_temp_df$month <- factor(avg_temp_df$month, levels = c("Jan", "Feb", "Mar","Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

avg_temp_df$season <- factor(avg_temp_df$season, levels = c("Winter", "Spring", "Summer", "Autumn"))

ggplot(avg_temp_df, aes(x = month, y = avg_temp, group = season, color = season)) +
  geom_point() +
  geom_line() +
  labs(title = "Monthly Temperature",
       x = "Month",
       y = "Temperature (Celsius)")
  
```

Above is a graph of the mean temperature in Celsisus of each month, colored by season, including lines also colored by season connecting the points by season.

### Part B - Similar plot with four sets of lines -  mean monthly Temp (in C), PM10, )3, and dewpoint

Produce a similar plot with four sets of lines: mean monthly temperature in celsius, mean montly O3, mean monthly PM10, and mean monthly dewpoint. Figure out a way to distinguish between the four lines. Again, make sure the plot is publication-ready (clean and clear labels). Use the plot to answer the following question:

\"Which of the four variables seems to have the least seasonal trend?\"

Create a df with averages needed:

```{r}
# Calculate the average temperature for each month
avg_temp_per_month <- tapply(nnmaps$tempC, nnmaps$month_numeric, mean)
avg_03_per_month <- tapply(nnmaps$o3, nnmaps$month_numeric, mean)
avg_PM10_per_month <- tapply(nnmaps$pm10, nnmaps$month_numeric, mean, na.rm = TRUE)
avg_dewpoint_per_month <- tapply(nnmaps$dewpoint, nnmaps$month_numeric, mean)

# Create a new data frame with average temperatures and seasons
avg_df <- data.frame(month_num = names(avg_temp_per_month), avg_temp = avg_temp_per_month, avg_o3 = avg_03_per_month, avg_pm10 = avg_PM10_per_month, avg_dew = avg_dewpoint_per_month)
avg_df$month <- month.abb[as.numeric(avg_df$month_num)]

avg_df <- avg_df %>%
  mutate(season = case_when(
  month %in% c("Jan", "Feb", "Mar") ~ "Winter",
  month %in% c("Apr", "May", "Jun") ~ "Spring",
  month %in% c("Jul", "Aug", "Sep") ~ "Summer",
  month %in% c("Oct", "Nov", "Dec") ~ "Autumn"))
avg_df
```

Graph, similar to part (a):

```{r}
# Order the "Month" factor
avg_df$month <- factor(avg_temp_df$month, levels = c("Jan", "Feb", "Mar","Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

avg_df$season <- factor(avg_temp_df$season, levels = c("Winter", "Spring", "Summer", "Autumn"))

ggplot(avg_df, aes(x = month, y = c(avg_temp, avg_o3, avg_pm10, avg_dew), group = season, color = season)) +
  geom_point() +
  geom_line() +
  labs(title = "Monthly Temperature",
       x = "Month",
       y = "Temperature (Celsius)")
```
